Computing Direct Sum Decompositions

### 2024-12-27 paper/main.tex
version submitted to arXiv

\title{Computing Direct Sum Decompositions}
\author{Devlin Mallory and Mahrud Sayrafi}
\begin{document}
\maketitle

@@

An accompanying implementation in Macaulay2 is available online via the GitHub repository \\
\centerline{
  \href{https://github.com/mahrud/DirectSummands}
  {\texttt{https://github.com/mahrud/DirectSummands}}.}

@@

This condition is polynomial in the coefficients of powers of $t$ in $\det(A_{\l_1,\dots,\l_r}-t I)$ and thus in the $\l_i$; to see this, note that
if we write
  $$ f:=\det(A_{\l_1,\dots,\l_r}-t I)=t^n  +t^{n-1}b_{n-1}+\dots +t b_1 +b_0, $$
then $f$
  has an $n$-fold root exactly when
  $ f,\d f/\d t,\dots, \d^{n-1} f/\d t^{n-1} $
  vanish simultaneously; the resultant of these $n$ polynomials in the $n$ variables $b_i$ gives polynomial conditions in the $b_i$ for this to occur.
%\devlin{TODO: this is $n+1$ equations, so make sure it's okay}
 In our setting, the $b_i$ are themselves polynomials in the $\l_i$, and thus we have obtained polynomial equations defining the locus where $A_{\l_1,\dots,\l_r}$ fails to have distinct roots, and thus the complement $U$ is open.
\end{proof}

@@

\begin{rem}
We note that it is known already by \cite[Proposition~15]{Atiyah} that an endomorphism of an indecomposable vector $E$ cannot have two distinct eigenvalues. The utility of the preceding lemma is in producing an explicit direct sum decomposition of $E$.
\end{rem}

@@

\begin{exa}[Frobenius pushforward on Grassmannians]
  Consider the Grassmannian $X = \mathrm{Gr(2,4)}$. We may work over the Cox ring $S$,
  %\[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \O(D)). \]
  which in this case coincides with the coordinate ring
  \[ S = \frac{k[p_{0,1},p_{0,2},p_{0,3},p_{1,2},p_{1,3},p_{2,3}]}{p_{1,2}p_{0,3}-p_{0,2}p_{1,3}+p_{0,1}p_{2,3}}. \]
  Then in characteristic $p=3$ we have:
  \[ F_*\O_X = \O \oplus \O(-1)^{44} \oplus \O(-2)^{20} \oplus A^4 \oplus B^4, \]
  where $A$ and $B$ are rank-2 indecomposable bundles (c.f.~\cite{RSVdB22}).
\end{exa}

@@

\begin{exa}[Syzygies over Artinian rings]
  In recent work suggested by examples calculated using our algorithm, \cite{CDE+24} studies the indecomposable summands of syzygy modules over a Golod ring $(R,m,k)$. In particular, they find previously unexpected recurring behavior, specifically that the syzygy modules of the residue field are direct sums of only three indecomposable modules: the residue field $k$, the maximal ideal $m$, and an additional module $N = \Hom_R(m, R)$.

>>>>

\title{Computing Direct Sum Decompositions}

\author{Devlin Mallory and Mahrud Sayrafi}

%% \author{Devlin Mallory}
%% \affiliation{Basque Center for Applied Mathematics, Bilbao, Basque Country, Spain}
%% \email{\href{mailto:dmallory@bcamath.org}{dmallory@bcamath.org}}

%% \author{Mahrud Sayrafi}
%% \affiliation{Max Planck Institute for Mathematics in the Sciences, Leipzig, Germany}
%% \email{\href{mailto:mahrud@mis.mpg.de}{mahrud@mis.mpg.de}}

%% \subjclass[2020]{16D70; 14F06, 13A35}

\begin{document}
\maketitle

@@

An accompanying implementation in Macaulay2 \cite{M2} is available via the GitHub repository \\
\centerline{
  \href{https://github.com/mahrud/DirectSummands}
  {\texttt{https://github.com/mahrud/DirectSummands}}.}

@@

This condition is polynomial in the coefficients of powers of $t$ in $\det(A_{\l_1,\dots,\l_r}-t I)$ and thus in the $\l_i$; to see this, note that
if we write
  $$ f:=\det(A_{\l_1,\dots,\l_r}-t I)=t^n  +t^{n-1}b_{n-1}+\dots +t b_1 +b_0, $$
then $f$
  has an $n$-fold root exactly when
  $ f,\d f/\d t,\dots, \d^{n-1} f/\d t^{n-1} $
  vanish simultaneously; the resultant of these $n$ polynomials in the $n$ variables $b_i$ gives polynomial conditions in the $b_i$ for this to occur.
 In our setting, the $b_i$ are themselves polynomials in the $\l_i$, and thus we have obtained polynomial equations defining the locus where $A_{\l_1,\dots,\l_r}$ fails to have distinct roots, and thus the complement $U$ is open.
\end{proof}

@@

\begin{rem}
We note that it is known already by \cite[Proposition~15]{Atiyah57} that an endomorphism of an indecomposable vector $E$ cannot have two distinct eigenvalues. The utility of the preceding lemma is in producing an explicit direct sum decomposition of $E$.
\end{rem}

@@

\begin{exa}[Frobenius pushforward on Grassmannians]
  % https://arxiv.org/abs/1901.10956
  Consider the Grassmannian $X = \mathrm{Gr(2,4)}$. We may work over the Cox ring $S$,
  %\[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \O(D)). \]
  which in this case coincides with the coordinate ring
  \[ S = \frac{k[p_{0,1},p_{0,2},p_{0,3},p_{1,2},p_{1,3},p_{2,3}]}{p_{1,2}p_{0,3}-p_{0,2}p_{1,3}+p_{0,1}p_{2,3}}. \]
  Then in characteristic $p=3$ we have:
  \[ F_*\O_X = \O \oplus \O(-1)^{44} \oplus \O(-2)^{20} \oplus A^4 \oplus B^4, \]
  where $A$ and $B$ are rank-2 indecomposable bundles (c.f.~\cite{RSVdB22}).
\end{exa}

@@

%\begin{exa}[Frobenius pushforward on invariant rings of finite groups]
  % https://arxiv.org/abs/2312.11786
%  \mahrud{TODO: add an invariant ring example from \cite{HS24}}
%\end{exa}

\begin{exa}[Syzygies over Artinian rings]
  % https://arxiv.org/abs/2208.05427
  In recent work suggested by examples calculated using our algorithm, \cite{CDE+24} studies the indecomposable summands of syzygy modules over a Golod ring $(R,m,k)$. In particular, they find previously unexpected recurring behavior, specifically that the syzygy modules of the residue field are direct sums of only three indecomposable modules: the residue field $k$, the maximal ideal $m$, and an additional module $N = \Hom_R(m, R)$.
